---
title: "CaliTestScoreAnalysis"
author: "Duc-Anh Nguyen"
date: "2025-02-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The covariates that we are interested in:
+ county: county of school (42counties).

+ lunch: percent qualifying for reduced-price lunch, is a continuous variable ranging from 0 to 100. A higher value of lunch means more students in the school come from low-income families. It serves as a proxy for socioeconomic status (SES) in *education research*.

+ income: the median household income in the school district’s area. It serves as an indicator of socioeconomic status (SES) at the *community level*. Higher income levels often correlate with better educational resources, parental involvement, and overall school performance.

+ expenditure: expenditure per student. Higher values indicate that the school district spends more per student on education, resources, and infrastructure.

+ read: average reading score(scale:0-1000).

+ math: average math score per school(scale:0-1000).

+ performance: average score of math and reading per school (scale:0-1000).
```{r }
library(AER)
library(tidyverse)
library(lme4)
data("CASchools")
CASchools |> colnames()
data <-
  CASchools[ ,c("county",
              "lunch",
              "expenditure",
              "income",
              "read",
              "math")]
data$performance <- (data$math + data$read)/2
data |> head()
dim(data)
skimr::skim(data)
length(data$county) == length(unique(data$county))
table(data$county)
ggplot(aes(county), data = data) + geom_histogram(stat="count")

data |> colnames()


ggplot(aes(lunch, performance), data = data) + geom_point()
ggplot(aes(county, performance), data = data) + geom_point()
ggplot(aes(expenditure, performance), data = data) + geom_point()
ggplot(aes(income, performance), data = data) + geom_point()
```
From the sample it seems that lunch has a negative correlation with the performance, meanwhile higher incomes tend to correlate with higher perfomance results
As results from same state might be somehow correlated, we fit a mixed model. 

```{r }
ri_model <- lmer(performance~lunch + expenditure + income + (1|county), data = data)
ri_model |> summary()


variances <- as.data.frame(VarCorr(ri_model))[,c(1,ncol(as.data.frame(VarCorr(ri_model))))]

(ICC <- (variances$sdcor[1]^2)/((variances$sdcor[1]^2)+ variances$sdcor[2]^2))

```
the marginal correlation between average perfomance scores of schools in the same county is around 22%

From the plot the relationship of the perfomance scores and incomes seems monotonic but non linear, let us fit splines to the model   
```{r }
library(mgcv)
gam_model <-
  gam(performance ~ s(income, bs = "ps", m = c(3, 2)) + lunch + expenditure, data = data)
gam_model |> summary()

plot(gam_model)
```

Now we want to model the joint distribution of the two average scores math and read at schools. We will use copula for this task.
```{r }

library(VineCopula)
library(rvinecopulib)

u_cas = cbind.data.frame(
  math = pnorm(data$math, mean = mean(data$math), sd = sd(data$math)),
  read = pnorm(data$read, mean = mean(data$read), sd = sd(data$read))
)

u_cas |> head()

cop_cas = bicop(data = u_cas,
                family_set = "onepar",
                selcrit = "bic")

summary(cop_cas)
plot(cop_cas)

```



```{r }



```

